---
title: "Genome Wide Association Studies (GWAS)"
author: "Flavio Lozano-Isla"
format:
  html:
    toc: true
    toc-location: left
    number-sections: true
    self-contained: true
editor_options: 
  chunk_output_type: console
execute:
  echo: false
  warning: false
---

# Project Setup

```{r}
#| label:  setup

source("https://inkaverse.com/docs.r")

cran <- c("devtools", "tidyverse", "openxlsx", "rMVP", "CMplot")
git <- c("hemstrow/snpR") 

suppressPackageStartupMessages({
  for (pkg in cran) { 
    if( !require(pkg, character.only = TRUE) ) {
      install.packages(pkg)
      library(pkg, character.only = TRUE)
    } 
  }
  
  for (pkg in git) {
    if( !require(sub(".*/", "", pkg), character.only = TRUE) ) {
      devtools::install_github(pkg, upgrade = TRUE)
      library(sub(".*/", "", pkg), character.only = TRUE)
    }
  }
}); remove(cran, git, pkg)

cat("Project: ", getwd(), "\n")
session_info()
```

# Reference

-   (article) https://www.nature.com/articles/ncomms1467
-   (44K SNP set) http://www.ricediversity.org/data/
-   (data) http://ricediversity.org/data/sets/44kgwas/
-   https://github.com/hemstrow/snpR/blob/master/vignettes/snpR_introduction.Rmd
-   https://doi.org/10.32614/CRAN.package.rMVP
-   https://www.cog-genomics.org/plink/2.0/

# Tranformar datos

```{bash}
#| eval: false

plink2 --ped sativas413.ped \
       --map sativas413.map \
       --export vcf \
       --outsativas413
```

install.packages("vcfR")
library(vcfR)

vcf <- read.vcfR("RiceDiversity/sativas413.vcf")




# Import data

```{r}
pheno <- read.xlsx("RiceDiversity/sativas413.xlsx")
geno <- snpR::read_vcf("RiceDiversity/sativas413.vcf")
info <- read.xlsx("RiceDiversity/sativas413info.xlsx") %>% 
  select(id = "GSOR.ID", spop = "Sub-population") %>% 
  tibble()


sample_meta <- geno %>% 
  names() %>% 
  gsub("_.*", "", .) %>% 
  enframe(value = "pop") %>% 
  bind_cols(., info)

sample.meta(geno)$spop <- sample_meta$spop

sample.meta(geno)
snp.meta(geno)
```

# PCA raw-data

```{r}
pca <- plot_clusters(geno
                     , facets = "spop"
                     , check_duplicates = T
                     , plot_type = "pca"
                     )

pca$plots$pca
```

# rMVP: Memory-efficient, Visualization-enhanced, and Parallel-accelerated Genome-Wide Association Study

## Import data

```{r}
sample_meta <- geno %>% 
  names() %>% 
  enframe(value = "Taxa") %>% 
  mutate(HybID = gsub("_.*", "", Taxa)) %>% 
  select(!name)
  
pheno <- read.xlsx("RiceDiversity/sativas413.xlsx") %>% 
  base::merge(., sample_meta) %>% 
  select(Taxa, everything()) %>% 
  select(!HybID) %>% 
  write_csv("RiceDiversity/sativas413.csv")

MVP.Data(fileVCF = "RiceDiversity/sativas413.vcf"
         , filePhe = "RiceDiversity/sativas413.csv"
         , sep.phe = ","
         , priority = "speed"
         , fileKin = FALSE
         , filePC = FALSE
         , out = "RiceDiversity/gwas"
         )
```

## Kinship

```{r}
MVP.Data.Kin(fileKin = TRUE
             , mvp_prefix = 'RiceDiversity/gwas'
             , out = 'RiceDiversity/gwas')
```

## Data input

```{r}
genotype <- attach.big.matrix("RiceDiversity/gwas.geno.desc")
phenotype <- read.table("RiceDiversity/gwas.phe", head = TRUE)
mapping <- read.table("RiceDiversity/gwas.geno.map", head = TRUE)
kinship <- attach.big.matrix("RiceDiversity/gwas.kin.desc")
```

## GWAS

```{r}
phenotype %>% names()

trait <- phenotype %>% 
  select(1, "Panicle.length")

imMVP <- MVP(
  phe = trait
  , geno = genotype
  , map = mapping
  , K = kinship
  , method = c("GLM", "MLM", "FarmCPU")
  , outpath = "RiceDiversity/gwas"
  , file.output = TRUE
  , file.type = "csv"
  , ncpus = 4
  )

gc()
```

## PCA

```{r}
genoPath <- "RiceDiversity/gwas.geno.desc"
genopca <- attach.big.matrix(genoPath)
pca <- MVP.PCA(M = genopca, cpu = 1)

MVP.PCAplot(PCA = pca
            , pch = 19
            , outpath = "RiceDiversity/gwas"
            , file.output = T
            , file.type = "jpg"
            )
```

## SNP density-plot

```{r}
MVP.Report(MVP = imMVP
           , plot.type="d"
           , col=c("darkgreen", "yellow", "red")
           , file.type="jpg", dpi=300
           , outpath = "RiceDiversity/gwas"
           )
```

## Phenotype distribution

```{r}
MVP.Hist(phe = trait
         , outpath = "RiceDiversity/gwas"
         , file.type="jpg", breakNum=18, dpi=300)
```

## Manhatan

```{r}
MVP.Report(imMVP
           , plot.type = "m"
           , multracks = TRUE
           , threshold = c(1e-6, 1e-4)
           , threshold.lty=c(1,2)
           , threshold.lwd=c(1,1)
           , threshold.col=c("black","grey")
           , amplify = TRUE
           , bin.size = 1e6
           , chr.den.col = c("darkgreen", "yellow", "red")
           , signal.col = c("red","green")
           , signal.cex = c(1,1)
           , file.type = "jpg", memo = "", dpi = 300
           , outpath = "RiceDiversity/gwas"
        )
```

## Q-Q plot

```{r}
MVP.Report(imMVP
           , plot.type = "q"
           , col = c("dodgerblue1", "olivedrab3", "darkgoldenrod1")
           , threshold = 1e6
           , signal.pch = 19
           , signal.cex = 1.5
           , signal.col = "red"
           , box = FALSE 
           , multracks = TRUE
           , file.type="jpg",memo="",dpi=300
           , outpath = "RiceDiversity/gwas"
           )
```
